Bee brain size is positively associated with urban lifestyles but not with positive population trends

### 2022-10-20 Manuscript/Ms.Rmd
Changing way of downloading data. occ_search() takes too long, now I use occ_download()

bibliography: 
  - references.bib
---

@@

Keywords:

@@

Pollination is a fundamental ecosystem service that connects the earth’s vegetation and human economy. Pollination services are performed by both managed and wild pollinators, mostly bees [@james2008; @winfree2011], both of which might be affected by a range of current and projected human-related environmental changes. There are indeed clear shreds of evidence of recent declines in pollinator populations [@bartomeus2019], and parallel declines in the plants that rely upon them [@potts2010; @scheper2014]. For example, in the UK, the third part of wild pollinator species has decreased from 1980 to 2013 [@powney2019], including six of all 16 non-parasitic bumblebees (Williams & Osborne, 2009). The decline of pollinators currently observed is projected to increase in the future (Dormann et al., 2008). Because most wild plant species (80%) and crop species (75%) are directly dependent on insect pollination for fruit and seed production (Klein et al., 2007; Ollerton et al., 2011), there is an increasing concern that the loss of pollinators from both natural communities and agricultural areas will have severe negative consequences for ecosystem functions and services (Ashman et al., 2004; Klein et al., 2007).
Although the reason why many pollinator are declining is insuficiently understood, a factor that is considered central is the loss and alteration of their habitats (Brown & Paxton, 2009; Natural Research Council, 2006). The destruction and fragmentation of natural habitats and their replacement by anthropogenic landscapes present new challenges for survival and reproduction of organisms, increasing the risk of extinction by negative population growth (Brown & Paxton, 2009; Natural Research Council, 2006). However, not all pollinators species are equally affected by habitat loss and alteration. In bees, for example, there are also “winners” —like Bombus terrestris (Rasmont et al., 2008)— that are able to thrive in human-altered environments, and there are also other species that although do not proliferate in such environments can at least maintain self-sustaining populations (Biesmeijer et al., 2006; Bartomeus et al., 2013; Reemer et al., 2012; Collado et al., 2019). Human-dominated habitats —notably cities— offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants (Cane et al., 2006; Winfree et al., 2007; Collado et al., 2019). Although variation across species in tolerance to habitat loss and alteration is likely to have important demographic consequences, whether this is the reason why some pollinators are declining while others are increasing is currently unknown. Hence, the ability to adapt to human-dominated new habitats and to exploit new opportunities have become important for species to thrive (Manenti et al., 2013). For species that tolerate human-related disturbances, human-dominated habitats —notably cities— offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants (Cane et al., 2006; Winfree et al., 2007; Collado et al., 2019). The question is then why only some species are able to enjoy the benefits of an urban life.In a novel environment, the chances to successfully survive and reproduce largely depends on cognition, that is, the processes involved in gathering, storing and reacting to environmental information. Cognitive limitations have thus help explaining the low success of many species in novel environments in general, and urban areas in particular. For example, reptiles, amphibians and birds that flourished after being introduced to new regions outside their native ranges tend to have larger brains (relative to their body size) than unsuccessful invaders (Sol et al., 2008; Sol et al., 2005; Amiel et al., 2011). In vertebrates, relative brain size is a good proxy of problem-solving and learning (e.g. see chapter 2 of this thesis for its relationship with learning), which can help individuals to cope with the many challenges on novel environments. However, similar evidence is lacking for insects, in part because advanced cognitive abilities were not considered possible with their tiny brains. More important, there was no unequivocal evidence that brain size was related to cognitive performance. However, we now know that bees and other insects have advanced cognition, being able for instance to use tools to resolve new problems and learn socially how to do so (Alem et al., 2016). There is in addition evidence that cognitive differences in the brain are reflected in differences in their brain size (Chapter 2).We used previously published data on bee forest and urban preference (Collado et al., 2019) as opposite points of habitat complexity, expecting to find bigger brain sizes for urban dwellers because a wider array of cognitive abilities are needed to inhabit these habitats. Species population trends were extracted from Bartomeus et al., 2013 and both datasets were crossed with a newly collected brain weight database. We wanted to explore if there is a relationship between brain size, habitat preference, and population trends. Hence, our questions were the following:Are brain sizes, as a proxy of behavioural plasticity, related to habitat preferences?Is population stability related to brain size as it has been seen in other taxa?Are habitat preferences related to population stability? Or alternatively, are urban dwellers populations increasing?

>>>>

bibliography: 
  - references.bib

header-includes:
- \usepackage{orcidlink}

---

\textbf{Miguel Á. Collado$^{1}$} \orcidlink{0000-0002-4216-317X} \textbf{Jose B. Lanuza$^{2}$} \orcidlink{0000-0002-0287-409X} \textbf{and Ignasi Bartomeus$^{2}$} \orcidlink{0000-0001-7893-4389}

Departamento de xxxxx, Universidad de Sevilla, Seville, Spain

@@

Keywords: XXXX

@@

Pollination is a fundamental ecosystem service that connects the earth’s vegetation and human economy. Pollination services are performed by both managed and wild pollinators, mostly bees [@james2008; @winfree2011], both of which might be affected by a range of current and projected human-related environmental changes. There are indeed clear shreds of evidence of recent declines in pollinator populations [@bartomeus2019], and parallel declines in the plants that rely upon them [@potts2010; @scheper2014]. For example, in the UK, the third part of wild pollinator species has decreased from 1980 to 2013 [@powney2019], including six of all 16 non-parasitic bumblebees [@williams2009]. The decline of pollinators currently observed is projected to increase in the future [@dormann2008]. Because most wild plant species (80%) and crop species (75%) are directly dependent on insect pollination for fruit and seed production [@klein2007; @ollerton2011], there is an increasing concern that the loss of pollinators from both natural communities and agricultural areas will have severe negative consequences for ecosystem functions and services [@ashman2004; @klein2007].

Although the reason why many pollinator are declining is insuficiently understood, a factor that is considered central is the loss and alteration of their habitats [@brown2009]. The destruction and fragmentation of natural habitats and their replacement by anthropogenic landscapes present new challenges for survival and reproduction of organisms, increasing the risk of extinction by negative population growth [@brown2009]. However, not all pollinators species are equally affected by habitat loss and alteration. In bees, for example, there are also “winners” —like Bombus terrestris [@rasmont2008]— that are able to thrive in human-altered environments, and there are also other species that although do not proliferate in such environments can at least maintain self-sustaining populations [@biesmeijer2006; @bartomeus2013; Reemer et al., 2012????; @collado2019].

Human-dominated habitats —notably cities— offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants [@cane2006; @winfree2011; @collado2019]. Although variation across species in tolerance to habitat loss and alteration is likely to have important demographic consequences, whether this is the reason why some pollinators are declining while others are increasing is currently unknown. Hence, the ability to adapt to human-dominated new habitats and to exploit new opportunities have become important for species to thrive [@manenti2013]. For species that tolerate human-related disturbances, human-dominated habitats —notably cities— offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants [@cane2006; @winfree2011; @collado2019]. The question is then why only some species are able to enjoy the benefits of an urban life.In a novel environment, the chances to successfully survive and reproduce largely depends on cognition, that is, the processes involved in gathering, storing and reacting to environmental information. Cognitive limitations have thus help explaining the low success of many species in novel environments in general, and urban areas in particular. For example, reptiles, amphibians and birds that flourished after being introduced to new regions outside their native ranges tend to have larger brains (relative to their body size) than unsuccessful invaders [@sol2008; @sol2005; @amiel2011]. In vertebrates, relative brain size is a good proxy of problem-solving and learning (e.g. see chapter 2 of this thesis for its relationship with learning), which can help individuals to cope with the many challenges on novel environments. However, similar evidence is lacking for insects, in part because advanced cognitive abilities were not considered possible with their tiny brains. More important, there was no unequivocal evidence that brain size was related to cognitive performance. However, we now know that bees and other insects have advanced cognition, being able for instance to use tools to resolve new problems and learn socially how to do so [@alem2016]. There is in addition evidence that cognitive differences in the brain are reflected in differences in their brain size (Chapter 2). We used previously published data on bee forest and urban preference [@collado2019] as opposite points of habitat complexity, expecting to find bigger brain sizes for urban dwellers because a wider array of cognitive abilities are needed to inhabit these habitats. Species population trends were extracted from @bartomeus2013 and both datasets were crossed with a newly collected brain weight database. We wanted to explore if there is a relationship between brain size, habitat preference, and population trends. Hence, our questions were the following:Are brain sizes, as a proxy of behavioural plasticity, related to habitat preferences?Is population stability related to brain size as it has been seen in other taxa?Are habitat preferences related to population stability? Or alternatively, are urban dwellers populations increasing?
